Bandit Multiclass Linear Classification: \\ Efficient Algorithms for the Separable Case

### 2019-01-23 ICML-submission/empirical-comparison.tex
Spell check
polished the experiment section

We implemented
Algorithm~\ref{algorithm:algorithm-for-strongly-linearly-separable-examples},
Algorithm~\ref{algorithm:kernelized} with
kernel~\eqref{equation:rational-kernel}, and the \textsc{Banditron} algorithm.
We generated strongly and weakly separable data sets with $K=3$ classes. The
data sets are shown in~\autoref{figure:strongly-and-weakly-separable-datasets}.
Although, the data sets are shown as two-dimensional, we used a third dimension,
the so-called \emph{constant feature}, i.e. a feature that for every example is
set to $1$. We added the constant feature so that
\autoref{figure:strongly-separable-dataset} is actually strongly lineraly
separable. For both data sets, the sample size is $T=5\times 10^6$ and the
margin is $\gamma=0.05$.

\begin{figure}[h]
\centering
\begin{subfigure}[b]{0.23\textwidth}
\captionsetup{justification=centering}
\begin{center}
\hspace*{-0.3cm} \includegraphics[width=1.15\textwidth, trim={0, 0cm, 0, 0}, clip]{figures/strong_points}
\caption{Strongly separable case}
\label{figure:strongly-separable-dataset}
\end{center}
\end{subfigure}
\hfill
\begin{subfigure}[b]{0.23\textwidth}
\captionsetup{justification=centering}
\centering
\hspace*{-0.3cm}  \includegraphics[width=1.15\textwidth, trim={0, 0cm, 0, 0}, clip]{figures/weak_points}
\caption{Weakly separable case}
\label{figure:weakly-separable-dataset}
\end{subfigure}
\vspace*{-0.2cm}
\caption{Strongly and weakly generated data sets with $K=3$ classes. All
examples lie in the unit ball of $\R^2$. Class 1 is depicted red. Classes 2 and
3 are depcited gree and blue, respectively. $80\%$ of the examples belong to class 1,
$10\%$ belong to class 2 and $10\%$ belong to class 3. Class 1
occupies the angle interval $[-15^\circ, 15^\circ]$, while classes 2 and 3
occupy angle intervals $[15^\circ, 180^\circ]$ and $[-180^\circ, -15^\circ]$
respectively. We removed points that lie within margin $\gamma=0.05$ of the
linear separator.}
\label{figure:strongly-and-weakly-separable-datasets}
\end{figure}

We evaluated all three the algorithms on the two data sets. Since all three
algorithms are randomized, we run each algorithm $20$ times. \textsc{Banditron}
has exploration rate parameter $\epsilon$ for which we tried values $\epsilon
\in \{0.02, 0.01, 0.005, 0.002, 0.001, 0.0005 \}$. The average cumulative number
of mistakes up to round $t$ as a function $t$ are shown in
Figures~\ref{figure:number-of-mistakes-strongly-separable-dataset}
and~\ref{figure:number-of-mistakes-weakly-separable-dataset}.

We can see there is a dilemma for \textsc{Banditron}. When the
exploration rate $\epsilon$ is large, it suffers from the cost of exploration.
When the exploration rate is small, it cannot adapt its model quickly enough.
This problem is exacerbated when a majority class occupies a small region, like
class 1. Striking a balance between the two extremes leads to the
$\Theta(\sqrt{T})$ number of mistakes. In contrast, our algorithm makes
significant progress on every mistake it makes, leading to a finite mistake
bound.

Appendix~\ref{section:supp-to-experiment} shows decision boundaries
that each of the algorithms learns.

\begin{figure}
\centering
\includegraphics[width=0.45\textwidth]{figures/strong3}
\caption{Strongly separable case. Cumulative number of mistakes of various algorithms.}
\label{figure:number-of-mistakes-strongly-separable-dataset}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=0.45\textwidth]{figures/weak3}
\caption{Weakly separable case. Cumulative number of mistakes of various algorithms.}
\label{figure:number-of-mistakes-weakly-separable-dataset}
\end{figure}

>>>>

In this section, we provide an empirical evaluation on our algorithms, verifying their
effectiveness on linearly separable datasets.
We generate strongly and weakly linearly separable datasets with $K=3$ classes in
$\R^3$, where the third dimension of every example is always equal to $1$
(i.e. a \emph{constant feature}).
See Figures~\ref{figure:strongly-separable-dataset} and~\ref{figure:weakly-separable-dataset}
 for visualizations
of the two datasets, along with detailed descriptions of the two data distributions.

%~\autoref{figure:strongly-and-weakly-separable-datasets}
%For both datasets, the sample size is $T=5\times 10^6$ and the
%margin is $\gamma=0.05$.
%Projections of the examples from the two datasets onto their first two coordinates
% are shown in
%We added a third dimension
%to every example that is equal to $1$. In other words, we added the so-called
%\emph{constant feature}. The constant feature is necessary so that
%\autoref{figure:strongly-separable-dataset} is actually strongly linearly
%separable.

\begin{figure}[h]
\centering
\begin{subfigure}[b]{0.23\textwidth}
\captionsetup{justification=centering}
\begin{center}
\hspace*{-0.3cm} \includegraphics[width=1.15\textwidth, trim={0, 0cm, 0, 0}, clip]{figures/strong_points}
\caption{Strongly separable case}
\label{figure:strongly-separable-dataset}
\end{center}
\end{subfigure}
\hfill
\begin{subfigure}[b]{0.23\textwidth}
\captionsetup{justification=centering}
\centering
\hspace*{-0.3cm}  \includegraphics[width=1.15\textwidth, trim={0, 0cm, 0, 0}, clip]{figures/weak_points}
\caption{Weakly separable case}
\label{figure:weakly-separable-dataset}
\end{subfigure}
\vspace*{-0.2cm}
\caption{Strongly and weakly separable datasets with $K=3$ classes. Here
 we show projections of the examples onto their first two coordinates, which
 lie in the unit ball of $\R^2$. Class 1 is depicted red. Classes 2 and
3 are depicted green and blue, respectively. $80\%$ of the examples belong to class 1,
$10\%$ belong to class 2 and $10\%$ belong to class 3. Class 1
occupies the angle interval $[-15^\circ, 15^\circ]$, while classes 2 and 3
occupy angle intervals $[15^\circ, 180^\circ]$ and $[-180^\circ, -15^\circ]$
respectively. The examples are strongly and weakly linearly separable with a margin of $\gamma=0.05$, respectively.}
\label{figure:strongly-and-weakly-separable-datasets}
\end{figure}

%We removed points that lie within margin $\gamma=0.05$ of the
%linear separator.

We implement
Algorithm~\ref{algorithm:algorithm-for-strongly-linearly-separable-examples},
Algorithm~\ref{algorithm:kernelized} with
rational kernel~\eqref{equation:rational-kernel}, and the \textsc{Banditron} algorithm.
We evaluate these algorithms on the above two datasets, with a time horizon of
$T=5\times 10^6$.
\textsc{Banditron}
has exploration rate parameter $\epsilon$, for which we tried values $\epsilon
\in \{0.02, 0.01, 0.005, 0.002, 0.001, 0.0005 \}$.
Since all three
algorithms are randomized, we run each algorithm $20$ times.
The average cumulative number
of mistakes up to round $t$ as a function of $t$ are shown in
Figures~\ref{figure:number-of-mistakes-strongly-separable-dataset}
and~\ref{figure:number-of-mistakes-weakly-separable-dataset}.

We can see that there is a tradeoff in the setting of the exploration rate $\epsilon$
for \textsc{Banditron}. For a large $\epsilon$, \textsc{Banditron} suffers from the cost of exploration, whereas for small $\epsilon$, its model cannot be updated quickly enough.
As expected, Algorithm~\ref{algorithm:algorithm-for-strongly-linearly-separable-examples} has a small number of mistakes in the strongly linearly separable setting, while having a large number
of mistakes in the weakly linearly separable setting, due to the limited representation power of linear classifiers.
In contrast,
Algorithm~\ref{algorithm:kernelized} with rational kernel has a small number of mistakes in both settings, exhibiting strong adaptivity guarantees. Appendix~\ref{section:supp-to-experiment} shows the decision boundaries
that each of the algorithms learns.

%This problem is exacerbated when a class of large proportion occupies a small region
%(e.g. class 1).
%In contrast, our algorithms make
%significant progress on every mistake it makes, leading to a finite mistake
%bound.
%Striking a balance between the two extremes leads to the
%$\Theta(\sqrt{T})$ number of mistakes.

\begin{figure}
\centering
\includegraphics[width=0.45\textwidth]{figures/strong3}
\caption{The cumulative number of mistakes versus the number of rounds,
in the strongly separable case (\autoref{figure:strongly-separable-dataset}) for various algorithms.}
\label{figure:number-of-mistakes-strongly-separable-dataset}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=0.45\textwidth]{figures/weak3}
\caption{The cumulative number of mistakes versus the number of rounds,
in the weakly separable case (\autoref{figure:weakly-separable-dataset}) for various algorithms.}
\label{figure:number-of-mistakes-weakly-separable-dataset}
\end{figure}

### 2019-01-23 ICML-submission/from-weak-separability-to-strong-separability.tex
polished the experiment section

We construct a positive definite kernel function $k(\cdot, \cdot)$ whose
corresponding feature map $\phi(\cdot)$ transforms any sequence of \emph{weakly}
linearly separable examples to \emph{strongly} linearly separable sequence of
examples. Specifically, $\phi$ has the property that if a set of labeled
examples in $\B(\zero,1)$ is weakly linearly separable with a margin $\gamma$, then
after applying $\phi$ the examples become strongly linearly separable with a
margin $\gamma'$ and their squared norms are bounded by $2$. The parameter $\gamma'$ is
a function of the old margin $\gamma$ and the number of classes $K$, and is
specified in \autoref{theorem:margin-transformation} below.

The kernel function $k:\B(\zero,1) \times \B(\zero,1) \to \R$ is defined as
\begin{equation}
\label{equation:rational-kernel}
k(x,x') = \frac{1}{1 - \frac{1}{2}\ip{x}{x'}_{\R^d}} \; .
\end{equation}
Note that kernel function $k(x,x')$ can be evaluated in $O(d)$ time.

>>>>

We use a positive definite kernel function $k(\cdot, \cdot)$, namely
\emph{rational kernel}~\citep{Shalev-Shwartz-Shamir-Sridharan-2011}, whose
corresponding feature map $\phi(\cdot)$ transforms any sequence of \emph{weakly}
linearly separable examples to \emph{strongly} linearly separable sequence of
examples. Specifically, $\phi$ has the property that if a set of labeled
examples in $\B(\zero,1)$ is weakly linearly separable with a margin $\gamma$, then
after applying $\phi$ the examples become strongly linearly separable with a
margin $\gamma'$ and their squared norms are bounded by $2$. The parameter $\gamma'$ is
a function of the old margin $\gamma$ and the number of classes $K$, and is
specified in \autoref{theorem:margin-transformation} below.

The rational kernel $k:\B(\zero,1) \times \B(\zero,1) \to \R$ is defined as
\begin{equation}
\label{equation:rational-kernel}
k(x,x') = \frac{1}{1 - \frac{1}{2}\ip{x}{x'}_{\R^d}} \; .
\end{equation}
Note that kernel function $k(x,x')$ can be evaluated in $O(d)$ time.

### 2019-01-23 ICML-submission/introduction.tex
polished the experiment section

In Appendix~\ref{section:nearest-neighbor-algorithm}, we propose and analyze a
very different algorithm for weakly separable data. The algorithm is based on
the obvious idea that two points that are close enough must have the same label.

Finally, we study two questions related to the computational and
information-theoretic hardness of the problem. Any algorithm for the bandit
setting collects information in the form of so called \emph{strongly labeled}
and \emph{weakly labeled} examples. Strongly labeled examples are those for
which we know the class label. Weakly labeled example is an example for which we
know that class label can be anything except for a particular one class. In
Appendix~\ref{section:np-hardness-of-weak-labeling-problem}, we show that the
offline problem of finding a linear multiclass separator consistent with a set
of strongly and weakly labeled examples is NP-hard. In
Appendix~\ref{section:mistake-lower-bound-for-ignorant-algorithms}, we prove a
lower bound on the number of mistakes of any algorithm that uses only
strongly-labeled examples and ignores weakly labeled examples.

>>>>

In Appendix~\ref{section:nearest-neighbor-algorithm}, we propose and analyze a
very different algorithm for weakly separable data. The algorithm is based on
the obvious idea that two points that are close enough must have the same label.
\chicheng{I wonder if the mistake bound of this algorithm is too large to be included
as ``our contributions'' - perhaps we should only mention it as a folklore/baseline?}

Finally, we study two questions related to the computational and
information-theoretic hardness of the problem. Any algorithm for the bandit
setting collects information in the form of so called \emph{strongly labeled}
and \emph{weakly labeled} examples. Strongly labeled examples are those for
which we know the class label. Weakly labeled example is an example for which we
know that class label can be anything except for a particular one class. In
Appendix~\ref{section:np-hardness-of-weak-labeling-problem}, we show that the
offline problem of finding a multiclass linear classifier consistent with a set
of strongly and weakly labeled examples is NP-hard. In
Appendix~\ref{section:mistake-lower-bound-for-ignorant-algorithms}, we prove a
lower bound on the number of mistakes of any algorithm that uses only
strongly-labeled examples and ignores weakly labeled examples.
